---
title: "Kaggle Titanic"
output: html_document
---

[Titanic dataset from Kaggle](https://www.kaggle.com/c/titanic)

```
VARIABLE DESCRIPTIONS:
survival        Survival
                (0 = No; 1 = Yes)
pclass          Passenger Class
                (1 = 1st; 2 = 2nd; 3 = 3rd)
name            Name
sex             Sex
age             Age
sibsp           Number of Siblings/Spouses Aboard
parch           Number of Parents/Children Aboard
ticket          Ticket Number
fare            Passenger Fare
cabin           Cabin
embarked        Port of Embarkation
                (C = Cherbourg; Q = Queenstown; S = Southampton)

SPECIAL NOTES:
Pclass is a proxy for socio-economic status (SES)
 1st ~ Upper; 2nd ~ Middle; 3rd ~ Lower

Age is in Years; Fractional if Age less than One (1)
 If the Age is Estimated, it is in the form xx.5

With respect to the family relation variables (i.e. sibsp and parch)
some relations were ignored.  The following are the definitions used
for sibsp and parch.

Sibling:  Brother, Sister, Stepbrother, or Stepsister of Passenger Aboard Titanic
Spouse:   Husband or Wife of Passenger Aboard Titanic (Mistresses and Fiances Ignored)
Parent:   Mother or Father of Passenger Aboard Titanic
Child:    Son, Daughter, Stepson, or Stepdaughter of Passenger Aboard Titanic

Other family relatives excluded from this study include cousins,
nephews/nieces, aunts/uncles, and in-laws.  Some children travelled
only with a nanny, therefore parch=0 for them.  As well, some
travelled with very close friends or neighbors in a village, however,
the definitions do not support such relations.
```

```{r, echo=FALSE, message=FALSE}
library(ggplot2)
library(GGally)
library(grid)
library(gridExtra)
library(scales)
library(corrplot)
library(pROC)

library(randomForest)
library(ROCR)

setwd("~/kaggle/titanic")

# data loading
train <- read.csv("data/train.csv")
test <- read.csv("data/test.csv")
```

## Explorative Analysis

### Overview of training data

```{r, echo=FALSE, message=FALSE}
str(train)
```

### Basic Data Preparation

* Convert 'Survived' and 'Pclass' to factors
* Extract Cabin Letters
* Remove PassengerID, Name, Ticket, Cabin

```{r, echo=FALSE}

# extract cabin letter from cabin field
cabin_letter <- function(vec) {
  f <- factor(gsub(" .*|[0-9]", "", vec), levels = c("", LETTERS[1:7], "T"))
  levels(f)[1] <- "?"
  return(f)
}

# convert to factors: Survived, Pclass
train$Survived <- as.factor(train$Survived)
for (column in c("Pclass")) {
  train[,column] <- as.factor(train[,column])
  test[,column]  <- as.factor(test[,column])
}

# extract cabin letter
dflist <- list(train = train, test = test)
dflist <- lapply(dflist, function(df) {
  df$CabinLetter <- cabin_letter(df$Cabin)
  df
})
train <- dflist$train
test <- dflist$train

# drop PassengerId, Name, Ticket, Cabin
train <- train[, !(names(train) %in% c('PassengerId', 'Name', 'Ticket', 'Cabin'))]
```

### Pairwise Feature Comparison

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#for (c in integerCols) numericDF[,c] <- jitter(numericDF[,c])
#plot(numericDF, col=as.factor(train$Survived))

gg <- ggpairs(data = train,
#        columns = c(numericCols, "Sex", "Survived"),
        upper = list(continuous = "smooth", combo = "box"),
        mapping = (ggplot2::aes(alpha = 0.1, color = Survived))
        )
gg + theme(text = element_text(size=6), axis.text = element_text(size=5), axis.title = element_text(size=6))
```

```{r, echo=FALSE, message=FALSE}
# get numeric features
numericCols <- names(train)[sapply(train, class) %in% c("numeric")]
integerCols <- names(train)[sapply(train, class) %in% c("integer")]

# remove 'Survived' and 'passengerId'
integerCols <- integerCols[!(integerCols %in% c("Survived", "PassengerId"))]
numericCols <- c(numericCols, integerCols)

cat("Numeric features:", paste0(numericCols, collapse=", "))
```

### Survival by Sex, Age, and Pclass
```{r, echo=FALSE, warning=FALSE}

gAge <- ggplot(train, aes(x=Survived, y=Age))
gAge <- gAge + geom_violin(adjust = 0.5, alpha = 0.3, width=0.5)
gAge <- gAge + geom_dotplot(binaxis='y', stackdir='center', dotsize=0.8, alpha= 0.4, binwidth = 1, position=position_jitter(width = 0.05), aes(fill=Sex, col=Sex))
gAge <- gAge + facet_grid(Pclass ~ Sex)
#gAge <- gAge + theme(legend.position = c(1, 0.3), legend.justification = c(1, 1), legend.text=element_text(size=6), legend.background = element_rect(fill=alpha('white', 0.5)))
gAge + coord_flip()

#gSex <- qplot(y = Survived, x = Sex, data = train, position="jitter", col = Sex, dotsize=0.5, alpha= 0.4) + facet_grid(. ~ Sex)
#grid.arrange(gAge + coord_flip(), gSex + ylab("") + theme(legend.position = "none"), ncol=2)
```


### Age and Fare Analysis

```{r, echo=FALSE, warning=FALSE}

# scatterhist plot based on https://gist.github.com/mkuhn/5434709
g_legend<-function(a.gplot){
    tmp <- ggplot_gtable(ggplot_build(a.gplot))
    leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
    legend <- tmp$grobs[[leg]]
    return(legend)
}

hide_legend <- theme(legend.position = "none")

# simple plots
ageFare <- qplot(data=train, Age, Fare, color=Pclass, xlim=c(0,80), ylim=c(0,520))
ageHist <- qplot(data=train, Age, binwidth=1, fill=Pclass, xlim=c(0,80)) +
  hide_legend + labs(x=NULL) +
  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank())

fareHist <- qplot(data=train, Fare, binwidth=10, fill=Pclass, xlim=c(0,520)) +
  hide_legend + labs(x=NULL) +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  coord_flip()
grid.arrange(ageHist, g_legend(ageFare), ageFare + theme(legend.position = "none"), fareHist,
             nrow=2, ncol=2, widths=c(3/4,1/4), heights=c(1/5,4/5))
```

### Categorial Feature Analysis

```{r, echo=FALSE}
gSurvived <- qplot(data=train, Survived, fill=Pclass, ylim=c(0,600)) + hide_legend
gSex      <- qplot(data=train, Sex,      fill=Pclass, ylim=c(0,600)) + hide_legend
gPclass   <- qplot(data=train, Pclass,   fill=Pclass, ylim=c(0,600))
gEmbarked <- qplot(data=train, as.factor(Embarked), fill=Pclass, xlab="Embarked", ylim=c(0,680)) + theme(legend.position = "none")
gSibSp    <- qplot(data=train, as.factor(SibSp), fill=Pclass, xlab="SibSp", ylim=c(0,680)) + theme(legend.position = "none")
gParch    <- qplot(data=train, as.factor(Parch), fill=Pclass, xlab="Parch", ylim=c(0,680))

grid.arrange(gSurvived, gSex, gPclass, gEmbarked, gSibSp, gParch, nrow=2, ncol=3)
```

```{r, echo=FALSE}
# pre-processing and clean-up

clean_data <- function(df) {
  df$Age[is.na(df$Age)] <- mean(df$Age, na.rm = TRUE) # fill NA with mean Age
  df$Fare[is.na(df$Fare)] <- mean(df$Fare[df$Pclass == 3]) # fill NA with mean Fare
  df$Embarked[df$Embarked == ""] <- "S" # fill empty Embarked with 'S' (most frequent)
  df$Embarked <- droplevels(df$Embarked)
  return(df)
}

# fill empty embarked cities as "S"
train[train$Embarked == "", ]$Embarked <- "S"
train$Embarked <- droplevels(train$Embarked)
#test[test$Embarked == "", ]$Embarked <- "S"

# cut Fares in groups
# train$FareGroup <- cut(train$Fare, breaks=hist(as.numeric(train$Fare), breaks="Scott", plot=FALSE)$breaks, include.lowest = TRUE) # too skewed
fareQuantiles <- quantile(train$Fare, probs=seq(0, 1, 0.1))
train$FareQuant <- cut(train$Fare, breaks=fareQuantiles, include.lowest = TRUE)
test$FareQuant  <- cut(test$Fare, breaks=fareQuantiles, include.lowest = TRUE)

#test[is.na(test$Fare),]$Fare <- mean(train[train$Pclass == 3,]$Fare)


# ensure identical levels for train and test
levels(train$Parch) <- levels(test$Parch) <- seq(0,9)

```

#### fixing missing age with random values

before fix:
```{r, echo=FALSE}
summary(train$Age) # 177 NA's

train$MissingAge <- is.na(train$Age)
```

after fix:
```{r, echo=FALSE}
# fix age
ageMin <- min(train$Age, na.rm = TRUE)
ageMax <- max(train$Age, na.rm = TRUE)
train$Age[is.na(train$Age)] <- as.integer(runif(sum(is.na(train$Age)), ageMin, ageMax))
test$Age[is.na(test$Age)] <- as.integer(runif(sum(is.na(test$Age)), ageMin, ageMax))

summary(train$Age) # 0 NA's
```

### Fare Distribution
```{r, echo=FALSE}
ggplot(train, aes(y = Fare, x = Pclass)) + geom_point(color="blue", alpha="0.2", position="jitter")
```

### Known Cabins in Pclass
```{r, echo=FALSE, warning=FALSE}
# plot Cabin Letters
ggplot(train, aes(y=Survived, x=CabinLetter, col = Pclass)) + geom_jitter() + facet_grid(Pclass ~ .)
```

### Correlation of Variables
```{r, echo=FALSE, warning=FALSE}
df <- with(train, data.frame(Pclass, Sex, Age, SibSp, Parch, Fare, FareQuant, Embarked, CabinLetter, MissingAge))
df[] <- lapply(df, as.integer)

df$FamilySize <- df$Parch + df$SibSp + 1
df <- df[, !(names(df) %in% c('CabinLetter', 'FareQuant', 'Pclass', "SibSp", "Parch"))]

corrplot(cor(df, method = "spearman"), order = "AOE")
```

### Prediction Models

#### Random Forest

Variable Importance:

```{r, echo=FALSE, warning=FALSE}
str(train)

fit <- randomForest(Survived ~ ., data = train, ntree=1000, importance=TRUE)
importance(fit)

par(mar=c(5,3,2,2)+0.1)
varImpPlot(fit, main = NULL) # or individually: varImpPlot(fit, type=1|2)
# plot(fit, log="y") # show errors

# confusion matrix
fit$confusion
```

```{r, echo=FALSE, warning=FALSE}
# prepare data frame for training
df <- with(train, data.frame(Survived, Pclass, Sex, Age, SibSp, Parch, Fare, FareQuant, Embarked, CabinLetter))
fit <- randomForest(Survived ~ ., data = df)

prob <- predict(fit, type = "prob")[,2]

# roc curve (ROCR)
pred <- prediction(prob, train$Survived)
perf <- performance(pred, "tpr", "fpr")
#plot(perf, avg= "threshold", colorize=T, lwd= 3, main= "With ROCR you can produce standard plots like ROC curves ...")
rocRF <- data.frame(fpr = unlist(perf@x.values),
                    tpr = unlist(perf@y.values),
                    alpha = unlist(perf@alpha.values))
#ggplot(rocRF, aes(x = fpr, y = tpr)) + geom_line() +
#  coord_fixed(ratio = 1) +
#  xlab("False Positive Rate") +
#  ylab("True Positive Rate") +
#  scale_x_continuous(breaks=seq(0, 1, 0.2)) + # Ticks from 0-1, every .2
#  scale_y_continuous(breaks=seq(0, 1, 0.2))   # Ticks from 0-1, every .2
```

### Logistic Regression

```{r, echo=FALSE}
train$Pclass <- as.numeric(train$Pclass)
train$SibSp <- as.numeric(train$SibSp)
train$Parch <- as.numeric(train$Parch)

logdf <- with(train, data.frame(Survived, Pclass, Sex, Age, SibSp, Parch, Fare, Embarked))

# normalize numeric columns
logdf[, numericCols] <- scale(logdf[, numericCols])

fit <- glm(Survived ~ ., data=logdf, family="binomial")
summary(fit)
#coef(fit)
#confint(fit)

prob <- predict(fit, type = "response")

# draw ROC curve (ROCR package)
pred <- prediction(prob, train$Survived)
perf <- performance(pred, "tpr", "fpr")
rocLR <- data.frame(fpr = unlist(perf@x.values),
                    tpr = unlist(perf@y.values),
                    alpha = unlist(perf@alpha.values))
# ggplot(rocLR, aes(x = fpr, y = tpr)) + geom_line() +
#   coord_fixed(ratio = 1) +
#   xlab("False Positive Rate") +
#   ylab("True Positive Rate") +
#   scale_x_continuous(breaks=seq(0, 1, 0.2)) + # Ticks from 0-1, every .2
#   scale_y_continuous(breaks=seq(0, 1, 0.2))   # Ticks from 0-1, every .2
```

```{r, echo=FALSE, warning=FALSE}
# combine both roc plots
rocRF$model <- "Random Forest"
rocLR$model <- "Logistic Regression"
rocDF <- rbind(rocRF, rocLR)
ggplot(rocDF, aes(x=fpr, y=tpr, col = model)) + geom_line() +
  coord_fixed(ratio = 1) +
  xlab("False Positive Rate") +
  ylab("True Positive Rate") +
  scale_x_continuous(breaks=seq(0, 1, 0.2)) + # Ticks from 0-1, every .2
  scale_y_continuous(breaks=seq(0, 1, 0.2))   # Ticks from 0-1, every .2
```

```{r, echo=FALSE, warning=FALSE}
export <- function(filename, result) {
  result <- cbind(test$PassengerId, as.data.frame(result))
  names(result) <- c("PassengerId", "Survived")
  result$Survived <- as.numeric(result$Survived) -1
  write.table(result, filename, sep=",", row.names=FALSE)
}

#pred <- predict(fit, test)
#export("predictionForest.csv", pred)
```
